Synthesizing Quotient Lenses

### 2018-03-12 papers/QuotientLenses/main.tex
A few responses and further comments

let bib_to_end : (lens in bibtex $\Leftrightarrow$ end_note) = preamble . label . author_lens . title_lens
\end{lstlisting}
\caption{An example bijective lens \cd{bib_to_end} between data
matching \cd{bibtex} and \cd{end_note} regular 
expressions, which are omitted for brevity.  
\bcp{And a readability suggestion:
define a named constant for a quote character.}}
\label{fig:example-lens}
\end{figure}

@@

In this subsection, we develop a QRE specification of \bibtex{}
records, introducing various QRE combinators along the way. 
Our first step in this process is to define a
whitespace format, which externally matches any non-zero number of whitespace
characters. It converts any such whitespace into a single space
character, its canonical form. 
We use the QRE \kw{project} primitive to define this
whitespace-normalizing QRE. \bcp{Why is it called
``project''?} \ksf{select seems like it might be a better name as we
are choosing one representative element from a set of possibilities.}

@@

\section{Quotient Regular Expressions}
A Quotient Regular Expression (or QRE) is a regular expression $R$ augmented
with syntax that expresses an equivalence relation on the language of $R$. This
section describes in further detail the set of QRE combinators that we described
in \cref{sec:example}.
\label{QRE}
\subsection{Syntax and Semantics of QREs}
Formally, the language of Quotient Regular Expressions (QREs) is given by the
following grammar:
\begin{align*}
Q := \; &R \sep \project{R}{s} \sep \squash{Q_1}{Q_2}{f} \sep
\perm{Q_1, \ldots, Q_n}{Q} \\
& | \; \normalize{R}{R'}{f} \sep Q_1 \semicolon Q_2 \sep Q_1 \cdot Q_2 \sep (Q_1
\sep Q_2) \sep Q^*,
\end{align*}
where $Q$ ranges over QREs, $R$ ranges over regular expressions, $f$ ranges over
functions between regular languages, and $s$ ranges over character strings.

Using the conventional notation that $\mathcal{L}(R)$ is the language accepted
by the regular expression $R$, then each QRE $Q$ enables us to express
\begin{enumerate}
  \item a regular expression $W(Q)$ (the ``whole'' of $Q$),
  \item an equivalence relation $\eqrel{Q}$ on $\mathcal{L}(W(Q))$,
  \item a regular expression $K(Q)$ (the ``kernel'' of $Q$)
  such that $\mathcal{L}(K(Q))$ forms a complete set of representatives for
  $\eqrel{Q}$, and
  \item a ``canonizing'' function\bcp{Why is it called ``canonizer'' rather
  than ``canonize''??}\sam{because \canonizer(Q) is a function}
  $\canonizer(Q):\mathcal{L}(W(Q)) \longrightarrow \mathcal{L}(K(Q))$ which
  given any $w \in \mathcal{L}(W(Q))$, computes $\canonizer(Q)(w)$ as the
  unique $k$ in $\mathcal{L}(K(Q))$ such that $k$ is equivalent to $w$ mod
  $\eqrel{Q}$
\end{enumerate}
Figures~\ref{fig:wk} gives the inductive definitions for the whole and
kernel languages $W(Q)$ and $K(Q)$ of a QRE.
\begin{figure}[t]
\centering
\[
\begin{array}{l@{\quad}l@{\quad}l}

@@

\item The remaining eight benchmarks were derived from datasets found at 
  data.gov. These datasets include census
  population data, wage comparison data, demographics statistics data and
  crime index data. We chose these datasets since they were stored in XML, CSV
  and RDF formats which are more easily specified using regular expressions than
  plain text or PDF documents.  \BCP{Besides differing formats, why did we
    include these eight new 
    ones?  Are they significantly different (more complex, exercising
    different features of the synthesis algorithm or QRE syntax,
    ...)?}\afm{Solomon, could you fill this out?}
    \sam{Well these data formats were the ones I was playing with before we got
    the Augeus and flashfill examples. I chose them simply because they had
    formats which could be expressed using regular expressions. They basically
    use the same QREs as the flashfill and augeus examples, mostly the project
    QRE. The fields in these files are occur in the same order in each record;
    we tried using the perm combinator anyways but the synthesis algorithm took
    too long. I do not think that they are significantly different from the
    flashfill and augues examples.}\afm{So would it be appropriate to say it was
    these were merely the first examples in the data.gov website whose languages
    are regular?}
\end{enumerate}

@@

Now we turn to related work in synthesis. Much of the research in synthesis
assumes that the synthesizer is provided with a collection of examples. Optician
and Optometrist differ in that they require that the programmer supplies both
examples {\em and} format descriptions in the form of regular expressions or
QREs, though we are far from the only ones to consider type-based synthesis.
There is a trade-off here.  On the one hand, a user must have some programming
expertise to write regular expression (or QRE) specifications, and it requires
some work. On the other hand, such specifications provide a great deal of
information to the synthesis system, which decreases the number of examples
needed (often to zero), makes the system scale well, and allows it to handle
large, complex formats.  By providing these format specifications, the 
synthesis engine does not have to both infer the format of the data as well as
the transformations on it, obviating the need to infer tricky formats like
those involving nested iterations. Furthermore, by focusing on bidirectional
transformations, we limit the space of synthesized functions to bijective ones,
reducing the search space, and the expressiveness of the search space.

>>>>

let bib_to_end : (lens in bibtex $\Leftrightarrow$ end_note) = preamble . label . author_lens . title_lens
\end{lstlisting}
\caption{An example bijective lens \cd{bib_to_end} between data
matching \cd{bibtex} and \cd{end_note} regular 
expressions, which are omitted for brevity.  
\iflater
\bcp{And a readability suggestion:
  define a named constant for a quote character.}\fi}
\label{fig:example-lens}
\end{figure}

@@

In this subsection, we develop a QRE specification of \bibtex{}
records, introducing various QRE combinators along the way. 
Our first step in this process is to define a
whitespace format, which externally matches any non-zero number of whitespace
characters. It converts any such whitespace into a single space
character, its canonical form. 
We use the QRE \kw{project} primitive to define this
whitespace-normalizing QRE. \bcp{Why is it called
``project''?} \ksf{select seems like it might be a better name as we
are choosing one representative element from a set of possibilities.}\bcp{Or
``collapse''?}

@@

\section{Quotient Regular Expressions}

\BCP{What became of my suggestion that we start with just ``normalize'' and
  then, later, introduce the other, more specific (and pragmatic, and
  incomplete) ones?}

A Quotient Regular Expression (or QRE) is a regular expression $R$ augmented
with syntax that expresses an equivalence relation on the language of $R$. This
section describes in further detail the set of QRE combinators that we described
in \cref{sec:example}.
\label{QRE}
\subsection{Syntax and Semantics of QREs}
Formally, the language of Quotient Regular Expressions (QREs) is given by the
following grammar:
\begin{align*}
Q := \; &R \sep \project{R}{s} \sep \squash{Q_1}{Q_2}{f} \sep
\perm{Q_1, \ldots, Q_n}{Q} \\
& | \; \normalize{R}{R'}{f} \sep Q_1 \semicolon Q_2 \sep Q_1 \cdot Q_2 \sep (Q_1
\sep Q_2) \sep Q^*,
\end{align*}
where $Q$ ranges over QREs, $R$ ranges over regular expressions, $f$ ranges over
functions between regular languages, and $s$ ranges over character strings.

Using the conventional notation that $\mathcal{L}(R)$ is the language accepted
by the regular expression $R$, then each QRE $Q$ enables us to express
\begin{enumerate}
  \item a regular expression $W(Q)$ (the ``whole'' of $Q$),
  \item an equivalence relation $\eqrel{Q}$ on $\mathcal{L}(W(Q))$,
  \item a regular expression $K(Q)$ (the ``kernel'' of $Q$)
  such that $\mathcal{L}(K(Q))$ forms a complete set of representatives for
  $\eqrel{Q}$, and
  \item a ``canonizing'' function\bcp{Why is it called ``canonizer'' rather
  than ``canonize''??}\sam{because \canonizer(Q) is a function}\bcp{My ear
  consistently finds it quite awkward.  I'd be happier with
  ``canonize'' (which describes an action).}
  $\canonizer(Q):\mathcal{L}(W(Q)) \longrightarrow \mathcal{L}(K(Q))$ which
  given any $w \in \mathcal{L}(W(Q))$, computes $\canonizer(Q)(w)$ as the
  unique $k$ in $\mathcal{L}(K(Q))$ such that $k$ is equivalent to $w$ mod
  $\eqrel{Q}$
\end{enumerate}
Figures~\ref{fig:wk} gives the inductive definitions for the whole and
kernel languages $W(Q)$ and $K(Q)$ of a QRE.
\begin{figure}[t]
\centering
\[
\begin{array}{l@{\quad}l@{\quad}l}

@@

\item The remaining eight benchmarks were derived from datasets found at 
  data.gov. These datasets include census
  population data, wage comparison data, demographics statistics data and
  crime index data. We chose these datasets since they were stored in XML, CSV
  and RDF formats which are more easily specified using regular expressions than
  plain text or PDF documents.  \BCP{Besides differing formats, why did we
    include these eight new 
    ones?  Are they significantly different (more complex, exercising
    different features of the synthesis algorithm or QRE syntax,
    ...)?}\afm{Solomon, could you fill this out?}
    \sam{Well these data formats were the ones I was playing with before we got
    the Augeus and flashfill examples. I chose them simply because they had
    formats which could be expressed using regular expressions. They basically
    use the same QREs as the flashfill and augeus examples, mostly the project
    QRE. The fields in these files are occur in the same order in each record;
    we tried using the perm combinator anyways but the synthesis algorithm took
    too long. I do not think that they are significantly different from the
    flashfill and augues examples.}\afm{So would it be appropriate to say it was
    these were merely the first examples in the data.gov website whose languages
    are regular?}\BCP{Or would anything be lost by just deleting them?  It
    seems like we're tying ourselves in knots at bit trying to explain how
    they relate to the others and what is interesting about them, when the
    answer seems to be they don't and nothing.}
\end{enumerate}

@@

Now we turn to related work in synthesis. Much of the research in synthesis
assumes that the synthesizer is provided with a collection of examples. Optician
and Optometrist\bcp{Sole use?!} differ in that they require that the
programmer supplies both 
examples {\em and} format descriptions in the form of regular expressions or
QREs, though we are far from the only ones to consider type-based synthesis.
There is a trade-off here.  On the one hand, a user must have some programming
expertise to write regular expression (or QRE) specifications, and it requires
some work. On the other hand, such specifications provide a great deal of
information to the synthesis system, which decreases the number of examples
needed (often to zero), makes the system scale well, and allows it to handle
large, complex formats.  By providing these format specifications, the 
synthesis engine does not have to both infer the format of the data as well as
the transformations on it, obviating the need to infer tricky formats like
those involving nested iterations. Furthermore, by focusing on bidirectional
transformations, we limit the space of synthesized functions to bijective ones,
reducing the search space, and the expressiveness of the search space.
